---
title: C++ Ternary Operator in R
author: Michael
date: '2023-01-14'
slug: c-ternary-operator-in-r
categories:
  - R
tags: []
---


Here's another fun but entirely useless exercise that shows some of the flexibility of R's infix operators.
C++ (and many other languages) have a ternary if-else operatory. 
You can write the following:

```c++
predicate ? value_if_true : value_if_false
```

"If Predicate then value_if_true, otherwise value_if_false".

In R, we have `ifelse`, or `dplyr::if_else`, and we can also put simple `if` on single lines:

```R
if (predicate) value_if_true else value_if_false
```

But if we want, we can abuse the ability to make infix functions as so.
First we define a two-argument function, `%?%` that takes a predicate on the left, and some expression on the right.


```{r}
`%?%` <- function(lhs, rhs) {
  rhs_prime <- substitute(rhs)

  list(payload = rhs_prime, predicate = lhs)
}
```

We have to `substitute()` the `rhs`, because we don't want R to evaluate that code straight away, we want to be able to make that decision once we've evaluated the predicate.

Our return value is a list with two elements.
The frozen `rhs` value as `payload` and the predicate value passed in as `lhs`.

Then we define another function, `%:%` that handles the `if` part:

```{r}
`%:%` <- function(lhs, rhs) {
  predicate <- lhs$predicate

  if (predicate) {
    eval(lhs$payload)
  } else {
    rhs
  }
}
```

Here, we just unpack the input from the `lhs` and decide which branch to evaluate based on the `predicate`.

In action:

```{r}
(4 < 10) %?% "lower" %:% "higher"
(4 > 10) %?% "lower" %:% "higher"
```
And because we used `substitute()` we only ever evaluate the value that matches the right branch.
So if one of the branches raises an error, but we never go down that branch, the expression still works:

```{r}
(4 > 10) %?% stop("Error never reached") %:% "everything is fine"
```

