---
title: R, Julia, multiple Dispatch
author: Michael
date: '2021-05-31'
slug: r-julia-multiple-dispatch
categories:
  - r
  - Julia
tags: []
---

# Julia

[Julia](https://julialang.org/) is a fascinating, relatively new programming language.
It is intended to be as simple to program in as Python, as fast as C, with the statistical abilities of R and the metaprogramming abilities of Lisp.
It's billed as a future major player in the scientific computer world, indeed it's the "Ju" in Jupyter.

One of Julia's key features is **multiple dispatch**. 
This is a feature that arises out of functions being defined by both their name *and* the types of their arguments.
Essentially, for a function, `my_func`, exactly what Julia will do depends on the arguments you pass in as well as the name of the function.
This `my_func(1)` is a totally separate function from `my_func("hello, there")`.

Julia is really interesting and I'd recommend checking the language out if you are interested.
A rather detailed introduction to multiple dispatch is [Stefan Karpinski's talk from Juliacon](https://www.youtube.com/watch?v=kc9HwsxE1OY&t=1699s),
but a more freidly place to get a feel for multiple dispatch is David Neuzerling's wonderful post [here](https://mdneuzerling.com/post/first-impressions-of-julia-from-an-r-user/).
I will be referring to this post later on, so I would recommend reading it before progressing here.


# R's S3 object system

R has a functional object oriented paradigm called S3.
S3 is R's most basic object oriented paradigm and it supports **single dispatch**.
This is the reason that we can call `summary` on a `data.frame` and get a certain type of response but we can also call `summary` on the result of a `lm()` call and get a totally different (and specialised) set of information.

This works because `summary` is a **generic** function.
Calling `summary` on an object of a particular class will cause R to look at what class the object is, and then look up a specific **method** of the generic that has been implemented for that class.

So when you call `summary` on a `data.frame`, R regonises that it has been given a `data.frame` and so passes your `data.frame` object to `summary.data.frame()`. 
When you call `summary` on `lm` objects, R passes your model to `summary.lm()`.
All this happens behind the scenes and you don't have to think about it.

S3 gives you quite a lot without asking much from you - it's a very beginner-friendly way to start working with R's object oriented frameworks, and Hadley Wickham's [Advanced R](https://adv-r.hadley.nz/s3.html) is a wonderful resource to learn more.

I like S3: I use it at work when I want to make an easy and consistent interface to code I write that abstracts particular implementation details.
Doing so makes my code easier to manage for me and easier for my colleagues to use.

But there is a little snag that makes S3 not as powerful as Julia's type system.
S3 cannot (easily?) do multiple dispatch, that is, we cannot define a function whose behaviour depends on the class of two or more of its arguments.
We can only match this behaviour to the first argument.

# R's S4 object system

R has a few other object systems, but the one I want to talk about in this post is S4.

I haven't used S4 much, and so later on in this post when I write some S4 code, don't read it as a recommendation:
I's more a "look what I did" rather than "this is how S4 should be done".
That said, S4 is on my list of things to learn in R and, again, Wickham's Advanced R is a useful starting point.

An interesting thing about S4 is that it supports **multiple** dispatch.

In the rest of this post, I will implement something similar to Neuzerling's Julia example.

# S4 Animals 

Neuzerling's example involved defining the way two animals can meet and interact.
They implemented both cat and dog types and a number of ways that these objects can interact all by overloading their `interact` function.
Again, if you haven't read their post, do so now or the motivation for the rest of mine may not be totally clear.

Our first step will be to define some classes.
We will start by defining a general `Animal` class, from which the rest of our classes will inherit:

```{r}
setClass("Animal",
         slots = c(
           name = "character"
         ))
```

S4 objects have 'slots' to put attributes.
We're going to be fairly basic in our implementation, so I'll just say that animals can have a name but not think about any other features.

Next we want to make a `Cat` and a `Dog` class:

```{r}
setClass("Cat", contains = "Animal")
setClass("Dog", contains = "Animal")
```

The `contains` argument to `setClass` defines the inheritance of both `Cat` and `Dog`.
This inheritance means that I don't need to specify that `Cat`s and `Dog`s also have names, because they get this by virtue of being `Animal`s.
Note that S4 allowed multiple inheritance too, but I'm not touching that right now.

We'll do one more thing now, which is to define an accessor function to get the `name` of an `Animal`.
Note that we *could* access the name using the `@` operator (which functions similarly to the `$` operator in, e.g. a named R list), but I understand that the `@` operator should really be reserved for programmers and not users.
So we implement a `name()` function like so:

```{r}
setGeneric("name", function(x) standardGeneric("name"))
setMethod("name", signature = "Animal", function(x) {x@name})
```

We first set `name` up as a generic and then define a specific method that will be applicable to all objects of class `Animal`.
Since `Cat`s and `Dog`s are a type of `Animal`, they get a `name` generic for free, just like the inherited having a name.

We can make specific instances of the animals like so:

```{r}
spot <- new("Cat", name = "Spot")
tom <- new("Cat", name = "Tom")

greg <- new("Dog", name = "Greg")
alice <- new("Dog", name = "Alice")
```

And we can get their `name`s like so:

```{r}
name(spot)
```

So far, so mundane: There's nothing here that we couldn't have done in S3.
Let's define what happens when these `Animal`s meet.

We first define the **generic** function `meet` which captures the general idea of objects (of any type) meeting.

```{r}
setGeneric("meet", function(x, y) standardGeneric("meet"))
```

Now, we have four different ways two `Animal` could meet:

- a `Cat` meets another `Cat`
- a `Dog` meets another `Dog`
- a `Cat` meets a `Dog`
- a `Dog` meets a `Cat`

But we should probably define a fallback/default method for `meet` that will work with any two `Animal`s, just in case we implement a new animal later on, but forget to implement the specific interactions:

```{r}
setMethod("meet",
          signature = c(x = "Animal", y = "Animal"),
          definition = function(x, y) {paste(name(x), "meets", name(y), sep = " ")})
```

Example:

```{r}
meet(spot, greg)
```

This call to `setMethod` includes the `signature` argument.
Here is where we define the classes of the arguments that will be passed in to the function defined by `definition`.
Here we state that when two `Animal`s meet, we supply a bland statement.

Next we must define the interactions we want to allow:

```{r}
setMethod("meet",
          signature = c(x = "Cat", y = "Cat"),
          definition = function(x, y) {paste(name(x), "hisses at", name(y))})

setMethod("meet",
          signature = c(x = "Dog", y = "Dog"),
          definition = function(x, y) {paste(name(x), "sniffs", name(y))})

setMethod("meet",
          signature = c(x = "Cat", y = "Dog"),
          definition = function(x, y) {paste(name(x), "bats", name(y))})

setMethod("meet",
          signature = c(x = "Dog", y = "Cat"),
          definition = function(x, y) {paste(name(x), "chases", name(y))})
```

- Two `Cat`s meeting hiss at each other
- Two `Dog`s meeting will sniff
- A `Cat` meeting a `Dog` will bat it
- A `Dog` meeting a `Cat` will chase it

Now if we run the same meeting as before, we get a different result, because R chooses the most specific method:

```{r}
meet(spot, greg)
```

And, we also have the following interactions:

```{r}
meet(spot, tom)
meet(greg, alice)
meet(alice, tom)
```

# Conclusion

I'm just dipping my feet into Julia, and I'm yet to move over completely from R.
Nevertheless, I already see the ways that multiple dispatch can be used to make more intuitive interfaces for users.

Likewise, I'm really new to S4 and I have much to learn about it, but it's nice to know that powerful features like multiple dispatch are available if I need them.

One word of warning would be that I don't know the performance costs of multiple dispatch in S4. 
I understand that there is no cost to Julia's multiple dispatch since it's a feature of the language, but I haven't investigated the costs in R.
I would imagine there could be a cost, and that would have to be investigated before committing to the implementation, but you've got to consider the mental costs of your users to struggle against a confusing interface vs the computational cost to your CPU to struggle against a particular object system.
Depending on what you're doing this trade off may come easily down on one side or the other or be about equal.


