---
title: When your code maybe works
author: Michael
date: '2022-08-03'
slug: when-your-code-maybe-works
categories:
  - r
  - functional programming
tags: []
---


## Functions to crash

R has a few ways to signal that something has gone wrong.
For example, there's the aptly-named `stop()` for when you need the program to stop completely:

```{r, error=TRUE}
stop("There was an error")
```

Or the newer `cli::cli_abort()`:

```{r, error=TRUE}
cli::cli_abort("There was an error")
```

As an aside, I like `cli::cli_abort()`, I think it's what powers all/most of the tidyverse errors (taking over from `rlang::abort()`).
It comes when {glue}-like syntax and fancier error prompts:

```{r, error=TRUE}
x <- 100
cli::cli_abort(c("There was an error because x had value: {x}",
                 "i" = "x cannot have value ",
                 "x" = "Don't do that again."))
```

But throwing an error like this will stop your program - if the code is being used interactively like how a lot of R will be used, that's probably the right thing to do.
If you write an informative error message that allows the user to pinpoint the issue, they can quickly fix it and get on their way. 
For example, say we try to access a column that doesn't exist using `dplyr::select()`:

```{r error=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(palmerpenguins)

penguins %>% select(mpg)
```

This immediately breaks and tells you exactly what's wrong.

But we have other options.
What if we wanted to handle functions that could fail in a bit more of an elegant way.

## `Option` or `Maybe`

Many stronger typed languages have things called `Option`^[e.g. Javascript] or `Maybe`^[e.g. Haskell]. 
These can be thought of as a box that _might_ contain a value, but could also contain nothing.

If you have a function that you'd really like to get an integer from usually, you may find when you write that function that there are cases where you won't get an integer back.
You could raise an error but if you want to be fancier, you could write your whole function to return a `Maybe Int`, that is, an integer wrapped in this `Maybe` box.

In the case that the function works you are provided with a box that has your integer in it.
This is often called a `Just`, as in `Just 4` or `Just 1003`.
The "`Just`" bit means that the function worked.

In the case that the function did not work, you'd get back a `Nothing`.
No error is thrown, but you know the function didn't work because of what you got back.

The key here is that both of these things, `Just 4` and `Nothing` are of the type `Maybe` in stronger typed languages.

R doesn't have a fantastic type system but we can pretend a bit using the S3 class system.

### `Maybe` in R

So we first need a way to make things of class `maybe`.
Recall that we want to be able to model two states:

1. a "success" state
2. a "failure" state

To do this, we will make two **constructor** functions for objects in our `maybe` class.

These are:
```{r}
just <- function(x) {
  output <- list(
    value = x,
    constructor = "just"
  )
  class(output) <- "maybe"
  output
}

nothing <- function(x) {
  output <- list(
    constructor = "nothing"
  )
  class(output) <- "maybe"
  output
}
```

Note the following things:

- In both instances we make an object with class `maybe`
- Both constructors are modelled as a list with a `constructor` element that holds whether this is a `just` or a `nothing`.
- The `nothing` has no `value` in its list.

We may as well add a nice `print` method too:

```{r}
print.maybe <- function(x) {
  if (x$constructor == "just") {
    cat("Just\n")
    print(x$value)
  } else if (x$constructor == "nothing") {
    cat("Nothing")
  }
}
```

This lets us show both states fairly neatly:

```{r}
just(4)

nothing()
```


### Functions that Fail Gracefully

So now, let's make some changes to the tidyverse to make functions that might fail, but do so by returning `maybe`s
Recall that earlier we couldn't `dplyr::select()` a column that didn't exist:

```{r, error=TRUE}
select(penguins, mpg)
```

What we want is a new function, which we will call `mselect()` that will return:

- `just(result)` if the function works, or
- `nothing()` if the function fails

I'm going to implement this as a simple wrapper over the {dplyr} function that just uses `tryCatch` to detect the error:

```{r}
mselect <- function(.data, ...) {
  tryCatch(just(dplyr::select(.data, ...)),
           error = function(e) nothing())
}
```


We're accepting the same arguments as `dplyr::select()` and trying them in that function.
The `tryCatch` function detects whether the `dplyr::select()` throws an error:

- if it doesn't, then `tryCatch` returns the result
- if it does, we execute the `error` function and return _that_ result.
  In this case, `nothing()`
  
In practice, this will look like:

```{r}
# works
mselect(penguins, species)

# doesn't work
mselect(penguins, mpg)
```

Fantastic, but now we can't pipe:

```{r, error=TRUE}
mselect(penguins, species) %>% 
  filter(species == "Adelie")
```

Because `dplyr::filter()` expects a `data.frame`, but got a `maybe` instead.

We could re-write all our functions to accept a `maybe` and then give back a `maybe`.
But then all our functions would have duplicated code to check whether the `maybe` is a `just` value or a `nothing`, then if it is a `just` value, we would need to extract that value and run the function.

If we were to implement that, it would mean there would be loads of duplication in our functions and the functions themselves wouldn't be doing a single thing well, they would be doing multiple things: Generally a bad way to design functions.

### More fun with pipes

But, we can do [interesting things with pipes](https://www.mpjon.es/2022/07/12/keeping-track-with-pipes/).
One of these will be to handle all the unwrapping and re-wrapping of our `maybe` values for us.

The full definition is:

```{r}
`%>>=%` <- function(lhs_prime, rhs) {
  
  if (!inherits(lhs_prime, "maybe")) {
    stop("type error")
  }
  
  if (lhs_prime$constructor == "nothing") {
    nothing()
  } else {
    # Regular pipe stuff
    rhs <- substitute(rhs)
    lhs <- substitute(lhs_prime$value)
    kind <- 1L
    env <- parent.frame()
    lazy <- TRUE
    .External2(magrittr:::magrittr_pipe)
  }
}
```

Things to note:

- We first check to make sure we're dealing with `maybe` values when we use this pipe.
  Fancier languages with fancier type systems would do this for us with their type-checking facilities as part of the language, but us type-paupers have to do it ourselves.
- Then we check if we are getting a `maybe` value that's a `nothing`.
  If we are, then whatever the function in the `rhs` is, we know we should be returning `nothing`.
  A function whose inputs have failed, must surely fail as well.
- Finally, if we have a `just` value, we pull out the `value` and then do regular pipe stuff from {magrittr}^[I still don't really know how it works yet].

To demonstrate, let's make another tidyverse function into a verison that returns a `maybe`.

```{r}
mfilter <- function(.data, ...) {
  tryCatch(just(dplyr::filter(.data, ...)),
           error = function(e) nothing())
}
```

It's to `dplyr::filter()` what `mselect()` was to `dplyr::select()`.

And now we can do things like:

```{r}
just(penguins) %>>=%
  mselect(species) %>>=%
  mfilter(species == "Chinstrap")
```

And our pipe-chain can fail at any points, but the `nothing` values will be seamlessly propagated through:

```{r}
just(penguins) %>>=%
  mselect(mpg) %>>=% # <-- fails here
  mfilter(species == "Chinstrap")
```

or

```{r}
just(penguins) %>>=%
  mselect(species) %>>=%
  mfilter(mpg == "Chinstrap") # <-- fails here
```

We have to start with `just(penguins)` rather than `penguins` as `%>>=%` expects a `maybe` on the left.
If this bothers you, you could make a new infix pipe that doesn't need a `maybe` on the left and use that as the first pipe.

### Simpler Piping

But we've had to individually redesign all our tidyverse functions.
I'd prefer not to have to copy and paste them all and make `m*` versions.

Luckily, we don't have to if we notice that `mselect()` and `mfilter()` both follow a similar pattern.
We can exploit this pattern by making a function factory:
A new function that will accept a tidyverse function as an argument and make a new version that returns a `maybe`.

```{r}
mwrap <- function(f) {
  function(...) {
    tryCatch(just(f(...)),
             error = function(e) nothing())
  }
}
```

Note how this function itself returns a function that looks like `mselect()` and `mfilter()` (if you blur your eyes a bit).

This means we can do pipelines like this:

```{r}
just(penguins) %>>=%
  mwrap(group_by)(species) %>>=%
  mwrap(summarise)(bill_length_mm = mean(bill_length_mm, na.rm=TRUE),
                   bill_depth_mm = mean(bill_depth_mm, na.rm = TRUE)) %>>=%
  mwrap(mutate)(bill_area = bill_length_mm * bill_depth_mm)
```


And if this fails at any point, we will get a `nothing` back.

## Conclusion

This implementation suffers a bit because R's type system isn't that strong.
In languages like Haskell, `Maybe`s are used a lot, and the `>>=` syntax is far more powerful than I have implemented here^[because `>>=` is defined for all monads, of which `Maybe` is only a single example].

Is this a useful thing? I don't know, that's up to you.

Is this an interesting thing? Yes, that's up to me, and yes it is.



